# Relation between Homeless and Unemployment 

Is there a relationship between homelessness and unemployment rate?

This repository contains all the R files that were produced to clean up mutiple csv files 
and a master csv file for visualization purpose; including identify the relationship using 
visualization technique. 

##### Direct Access

1. [Avg-work file](https://github.com/linhly2012/FinalProjectSOC225/blob/master/Final%20Project%20225/avg-work.R) 
2. [Homeless_unemploy_clean file](https://github.com/linhly2012/FinalProjectSOC225/blob/master/Final%20Project%20225/homeless_unemploy_clean.R)
3. [Homeless_work file](https://github.com/linhly2012/FinalProjectSOC225/blob/master/Final%20Project%20225/homeless_work.R)
4. [Joint_data_work file](https://github.com/linhly2012/FinalProjectSOC225/blob/master/Final%20Project%20225/joint_data_work.R)
5. [Pop-work file](https://github.com/linhly2012/FinalProjectSOC225/blob/master/Final%20Project%20225/pop-work.R)
6. [Unemployment_work file](https://github.com/linhly2012/FinalProjectSOC225/blob/master/Final%20Project%20225/unemployment_work.R)

```{r package_options, include=FALSE}
#Not including library packages loading when knit the file 
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)
library(tidyverse)
library(dplyr)
library(knitr)
```

##### Overview of the master data

```{r, echo = FALSE}
data <- read.csv('~/Desktop/FinalProjectSOC225/Final Project 225/clean_output/data.csv')
kable(data %>% select(-Estimate.Date) %>% rename("est_pop" = "Estimate") %>% head(5))
```



