---
title: "SOC 225 LAB Homework 1"
author: "Linh Ly"
output:
  html_document:
    df_print: paged
---
Write all code in the chunks provided!

Remember to unzip to a real directory before running everything!

## Problem 1: Import Data

1.1 Import the tidyverse
```{r}
library('tidyverse')
```

1.2 Read the airbnb file: data/airbnb.csv

```{r}
airbnb_data <- read_csv('airbnb.csv')

```

1.3 Output the column names using the function colnames()

```{r}
colnames(airbnb_data)
```


## Problem 2: Research Question (You don't need code for this question)
*For this problem you'll answer some questions to help explore your interests in data science. It's worth spending some time thinking about your interests and what you want to do as these questions will grow into your final poster project. That said, these are questions that you're interested in. They don't have to be things that you know the answer to and still less new areas of study.*  
  
*However, problem 3 asks you to come up with a 'big data' dataset that you think you might use to answer your question. If you're new to R or not sure about what to do, I encourage you to use the Airbnb data that we'll be using in class. In that case, make sure that your answers to problem 2 relate to the airbnb data.*

2.1: What are some areas of interest for you within sociology, big data, and computational social science?
- Homelessness, crime rate, income and education rate. 

2.2: Provide a link to a dataset which you think intersects with one of your interests. Explain the connection. You can find datasets by doing a google search or by looking here:http://hadoopilluminated.com/hadoop_illuminated/Public_Bigdata_Sets.html or here: https://www.kaggle.com/

Homelessness data with population data(2007 - 2016): 
https://www.kaggle.com/adamschroeder/homelessness/data

Unemployment rate: 
https://www.kaggle.com/jayrav13/unemployment-by-county-us

- Another dataset that I think intersects with my interest is unemployment rate. Because I wanted to see if unemployment rate has some sort of influence in the homelessness rate as the change of unemployment rate of each year.   

If you're using the airbnb set, explain how it connects to your interests.

2.3 Write a preliminary version of a research question that addresses your interest and points to how you might use the data set you've selected.

- Does the unemployment rate has influence on homelessness?

## Problem 3: Import Data and Identify Variables

3.1: Import your data in to R and output the column names
```{r}
#combine the unemployment to homelessness based on the county # 
homeless <- read.csv('data/homelessness/2007-2016-Homelessnewss-USA.csv')
unemployment <- read.csv('data/unemployment-by-county-us/output.csv')
```

3.2: Examine the variables in your data set. Which variables seem like they would be useful for you. Summarize those variables using summary() and indicate the types of data in those variables.

```{r}
colnames(homeless)
colnames(unemployment)
```
- The variables seem like they would be very useful is the "Count" values in homeless data and "Rate" in unemployment dataframe, as these two values will show the change in unemployment rate and homeless rate throughout times. 



