---
title: "Relationship between Unemployment and Homelessness"
author: "Linh Ly"
output: html_document
---
```{r package_options, include = FALSE, echo=FALSE}
#Not including library packages loading when knit the file 
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(knitr)
```

####Focus 

According to [U.S.News](https://www.usnews.com/news/us/articles/2017-12-06/us-homeless-count-rises-pushed-by-crisis-on-the-west-coast), homeless population in the nation begin to rise after a while (ever since 2010). The rise of homelessness is increasing fast especially in West Coast cities. It is important for us to understand the root grow of homeless and unemployment is one of the factor that can potentially contribute to an individual chance become homeless. The problem this research aim to understand: _"Is there a relationship between unemployment and homelessness ?"_

####Links/Sources of the data 

* Population: 
    + 2000 to 2010: https://www.census.gov/data/tables/time-series/demo/popest/intercensal-2000-2010-counties.html
    + 2010 to 2016: https://www.census.gov/data/tables/2016/demo/popest/state-total.html
* State Region / Division: https://github.com/cphalpert/census-regions/blob/master/us%20census%20bureau%20regions%20and%20divisions.csv
* Homelessness data(2007 - 2016): https://www.kaggle.com/adamschroeder/homelessness/data
* Unemployment rate: https://www.kaggle.com/jayrav13/unemployment-by-county-us

####Method 

- Using the provided sources, a master data sheet will put together to understand the change of unemployment rate and homelessness over time during the course of 2007 to 2016 after the data cleansing process. 

####Total Homelessness Sum
```{r homeless, echo= FALSE}
homeless <- read.csv('data/homelessness/2007-2016-Homelessnewss-USA.csv')
#note: the count value is not the rate. 
#cleaning text - change the year to actual year in numeric bc it is in 1/1/YYYY format
homeless$Year <- as.numeric(gsub("1/1/", "",as.character(homeless$Year))) 

#global variable - because this will be use more often through out the file
years <- sort(as.array(unique(homeless$Year)))

#Looking the data measure at "Total Homeless" (t.h) level-----------------------------------------
#calculate the sum of the totalhomeless count by year and state; and droping 
#these states from total homeless count because the unemployment does not has 
#data about these states. 
totalHomeless_sum <- homeless %>% filter(Measures == 'Total Homeless') %>% arrange(State) %>%
  group_by(Year, State) %>%
  summarise(total.homeless_sum = sum(as.numeric(gsub(",","",Count)))) %>%
  filter(!State %in% c("AK", "DC", "FL", "GA", "GU", "PR", "VI"))

kable(totalHomeless_sum %>% head(5))
```

####Unemployment Rate Sum 

```{r unemployment, echo = FALSE}
#load in data-------------------------------------------------
unemployment <- read.csv('data/unemployment-by-county-us/output.csv')

#filter the data down to january level to match with the way homeless data was structure
unemploy_2007.2016 <- unemployment %>% filter(Year >= 2007, Month == 'January')

#the function will take it a state name and filter the dataframe to only that state leve
#and then loops through each year to find the sum of those year in one dataframe for 
#for one state (of the county level)
addSumRateState <- function(name) {
  df <- unemploy_2007.2016 %>% filter(State == name)
  res <- c()
  for(i in years) {
    year_df <- df %>% filter(Year == i)
    res <- c(res,sum(year_df$Rate)) 
  }
  new_df <- data.frame(Year = years, State = name, 
                       unemploy.rate_sum = res)
  return(new_df)
}

#return a dataframe from 2007 to 2016 of each state unemployrate by year. 
unemployRate_sum <- lapply(sort(unique(unemploy_2007.2016$State)), FUN=addSumRateState) %>% bind_rows()

kable(unemployRate_sum %>% head(5))
```

#### Homeless and Unemploy Rate Sum 
```{r, echo = FALSE}
#the reason a state-code data frame was create to make sure when merge 
#homeless data and unemploy together, the data will go according to the state
#since the way the data structure is completely different 
state_code <- data.frame(state.code = state.abb, state.nam = state.name) %>% 
  filter(!state.abb %in% c("FL", "AK", "GA"),
         !state.name %in% c("Florida", "Alaska", "Georgia"))

totalHomeless_sum <- totalHomeless_sum %>% rename(state.code = State)
#merge homeless data with state code first
jointed_homeless_unemploy <- merge(x= totalHomeless_sum, y= state_code, all.x=TRUE)
#then merge unemploy after it - to prevent error 
unemployRate_sum <- unemployRate_sum %>% rename(state.nam = State) #for merging purpose
jointed_homeless_unemploy <- merge(x = jointed_homeless_unemploy, y = unemployRate_sum) %>% arrange(state.nam)

#merge df UNEMPLOY + HOMELESS + STATE CODE/NAME - to csv for later purpose
# write.csv(full_data, 
#           "~/Desktop/soc225/Final Project 225/clean_output/jointed.csv", 
#           row.names =FALSE)

kable(jointed_homeless_unemploy %>% head(5))
```

#### Homeless and Unemploy Rate Sum And Average

```{r finding average, echo = FALSE}
#calculating avg of homeless and unemploy data
avg <- jointed_homeless_unemploy %>% group_by(state.nam, state.code) %>% 
  mutate(mean.homeless.state = mean(total.homeless_sum),
         mean.unemploy.state = mean(unemploy.rate_sum)) %>% rename(state = state.nam)
kable(avg %>% rename("state code" = "state.code",
                      "homeless sum" = "total.homeless_sum", 
                      "unemploy sum" = "unemploy.rate_sum", 
                      "homeless mean" = "mean.homeless.state", 
                      "unemploy mean" = "mean.unemploy.state") %>% head(5))
```

#### Population Estimate from 2007 to 2016 

```{r population, echo = FALSE}
#load data 
state_pop <- read.csv('data/state-pop-10-16/state_est_10-16clean.csv')
county_pop <- read.csv('data/county-00-10/clean-county-00-10.csv')

#data preference note. 
#T population on county file in 2010 is different in county population and state file. 
#I chose state file because it seem like it was more updated since it was follow after 
#from 2010 to 2016 and for the county file it was from 2000 - 2010 
#---
#clean county population data, dropped 2001 - 2006 and 2010, because the data
#with homeless and unemployment measure go from 2007 to 2016; also due to the way
#the population data is gathered. Dropped city column because we only look at 
#state level (since the file did included total population of state from 2000 - 2010) . 
c_county <- county_pop %>% select(-c(SUMLEV, REGION, DIVISION, STATE, COUNTY,
                                     ESTIMATESBASE2000, CENSUS2010POP, EST_JUL_2000, 
                                     EST_JUL_2001, EST_JUL_2002, EST_JUL_2003, 
                                     EST_JUL_2004, EST_JUL_2005, EST_JUL_2006, EST_JUL_2010)
                                      ) %>% rename(state = STNAME, city = CTYNAME
                                                   ) %>% filter(city %in% state) %>% select(-city)

#filtered out any state data that was repeated more than one time. 
c_county <- c_county[!duplicated(c_county$state), ]

#dropped Puerto Rico data due to the structure of homeless and unemployment data frames. 
c_state <- state_pop %>% rename(state = GEO.display.label) %>% filter(state != "Puerto Rico")

#merging population measurement together from 2006 - 2009, and 2010 - 2016 from two different 
#data frame. 
pop_data <-merge(x = c_county , y =c_state) %>% select(-c(GEO.id, GEO.id2)) %>% 
  filter(!state %in% c("Alaska", "District of Columbia", "Florida", "Georgia"))

pop <- pop_data
colnames(pop) <- gsub("EST_JUL_", tolower("EST JUL "),colnames(pop))
kable(pop %>% head(5))
```

#### Master Data without Region and Division Code 

```{r master data, echo = FALSE}
#join AVG, UNEMPLOY, HOMELESS, POPULATION, YEAR, STATE data frame together
x <- suppressWarnings(suppressMessages(left_join(avg, pop_data, by = "state")))
#restructure 
master_data <- x %>% 
  gather(key = "Estimate Date", value = "Estimate", EST_JUL_2007:EST_JUL_2016) %>% 
  filter(Year == as.numeric(gsub("EST_JUL_", "",`Estimate Date`))) %>% 
  mutate(homeless_proportion = total.homeless_sum / Estimate, 
         unemploy_proportion = unemploy.rate_sum/ Estimate) %>%
  select(-c("Estimate Date")) %>% rename(est_pop = Estimate)

#for easy access later
colnames(master_data) <- tolower(colnames(master_data))

master <- master_data #for display purpose
colnames(master) <- gsub("[[:punct:]]+", " ",colnames(master))
kable(master %>% head(5))
```

#### Master Data with Region and Division Code 

```{r full master data, echo = FALSE}
region <- read.csv('census-regions-master/us census bureau regions and divisions.csv')
#View(region)

colnames(region) <- tolower(colnames(region))

join <- suppressWarnings(suppressMessages(master_data %>% left_join(region, by = "state")))
#View(join)
#restructure
join <- join %>% select(year, division, region, state, state.code.x, est_pop, 
                        total.homeless_sum, unemploy.rate_sum, 
                        mean.homeless.state, mean.unemploy.state, 
                        homeless_proportion, unemploy_proportion) %>% rename(state.code = state.code.x)

d_join <- join #for display purpose
colnames(d_join) <- gsub("[[:punct:]]+", " ",colnames(d_join))
kable(d_join %>% head(5))
```

####Summary statistics

```{r sum stat, echo = FALSE}
#using a dummy variable for displaying purpose
j <- join
colnames(j) <- gsub("[[:punct:]]+", " ",colnames(j))
kable(summary(j %>% ungroup() %>% select(-c(year, division, region, state, c("state code")))))
```

####Visualization
```{r graph 1, echo = FALSE }
ggplot(join %>% filter(between(year, 2010, 2016)), 
       aes(x=unemploy.rate_sum, y= total.homeless_sum, colour = region)) +
  geom_point() + 
  facet_wrap(~region, scales ="free") + 
  geom_smooth(fullrange = TRUE, method = "lm") + 
  labs(title = "Unemployment Rate and Homelessness of US Regions (2010 to 2016)") + 
  xlab("Unemployment Rate") +
  ylab("Homelessness") +
  theme(axis.title.y = element_text(margin = margin(r = 20)), 
        axis.title.x = element_text(margin = margin(t = 10)), 
        strip.text.x = element_blank())
```

- Figure 1: The measurement of total homelessness (that do not belong to specific category) in relation to the unemployment rate in 2010 to 2016 in 4 different regions of US. 

```{r graph 2, echo = FALSE}
ggplot(join %>% filter(between(year, 2010, 2016), 
                       state.code %in% c("CA", "WA", "TX", "MD", "NY", "MA", "WI", "MO")), 
       aes(unemploy_proportion, homeless_proportion, colour = state)) +
  geom_point() + 
  labs(title = "Relationship between Unemployment and Homeless Proportion (2010 to 2016)") + 
  xlab("Unemployment Proportion") +
  ylab("Homeless Proportion")
```

- Figure 2: The relation between homeless and unemployment proportion to the population of (selected) each state in 2010 to 2016 in 4 different US regions (8 states with 2 states from each region). 

#####Figure 1 and 2: 

######Connection of data  

- From figure 1, as the unemployment rate increase over time, the homelessness measure is being affected by it positively, especially for Midwest, South and West regions of the US. Using linear model, it also displayed that unemployment rate has influence of homelessness, especially the line also went through some plots on the graph. However, the Northeast region plot is questionable because it is pretty clustered together during early time and afterward, the number of unemployment rate is high but the homelessness either under the linear model or above. This is questionable, because there were at lease three states data for these two measurements are not included.  

- Looking into figure 2, it displayed that the higher the unemployment proportion, the lower the homeless proportion; the lower the unemployment proportion, the higher the homeless proportion. This figured was plotted by a selection of 8 states with 2 states from each regions. In a way this figure also raised a question, because if this was a positive correlated impact, then it is expected to see high unemployment proportion, high homelessness proportion like the relationship was shown in figure 1 between unemployment and homelessness. One factor that could cause this happen is the population of the states. Which make sense, for states like New York, California, or Washington, their populations has been growing for the last couple years because the larger the number divided to a large  number, the answer is smaller. 


```{r graph 3, echo = FALSE}
ggplot(join %>% filter(state.code %in% c("CA", "ID", "WA", "UT"))) + 
  geom_line(aes(unemploy.rate_sum, total.homeless_sum, colour = state)) +
  facet_wrap(~state, scale = "free") + 
  labs(title = "Unemployment Rate v. Homelessness in West Coast (2007 to 2016)") + 
  xlab(" Unemployment Rate") +
  ylab(" Total Homelessness") + 
  theme(strip.text.x = element_blank())

```

- Figure 3: The change of Unemployment Rate over time during the last 9 years of West Coast states - California, Oregon, Utah, and Washington. 

######Connection of data
- Using line graphs to understand the relationship between these two variables (unemployment and homelessness), this figures show that as the unemployment rate increase over time, the homelessness measure is also increasing, especially in popular state like California or Washington, which support the argument that there is a positive relationship from figure 1. 

####Dicussion - for all figures: 

######Limitations

- A limitation that these data restricted to was several states were dropped from the data frame and no explanation was provided. This can potentially skewed the unemployment rate and homelessness graph, since it factor goes by regions of US. Another limition is the way of population was constructed. The census releases population data from 2000 to 2010, and 2010 to 2016 in two different format, especially the estimate population of 2010 in these two population files are not the same - this can affect the graph outliers as well. The way the data produced not only can influenced the visualization due to outliers difference, but it will impacted the way people will interpreted the visualization as well. 

######Confidence

- The confidence in these findings are about 50% because the limitations of these data provided and the way the data was constructed. Additionally, it is hard to tell if the unemployment has impacted to individual homelessness, because the homelessness data that was accessible only available from 2007 to 2016. Even if the unemployment data included from 19XX to 2016, it is impossible to see the relationship. It will be easier, if there is an access to data of homeless from 2006 to 2016 or from 2000 to 2016. Also, the data from 2007 to 2009 were not graphed, due to the number of unemployment going on between 2007 to 2009 was high (outliers) with the economic fall during that time, it clustered any data that is being plotted after 2009. Including the way data constructed, I would assume that not every single locations or states (since one of the data did not have any values for at least three states) are being surveyed due to the geographies issues or they preferred not to be surveyed. 

######Ethical concern 

* For: 

    + Homeless data: 
        + it did not specify how these data was collected but only mentioned that this is a Point-in-Times estimates of homelessness. Although, the raw-data structured is well-formatted in a way of anonymity, but still can be concerning because it seem like they were using housing related data, which we don't know if the people who participated in these housing-related-data process know that this is being public. At the same time, I feel like you can potentially dig deeper who was involved with these data (especially in states or counties that has small population) because the county code was provided and specific homeless groups were labeled.

    + Unemployment data: 
        + Although the author of this data mention how it was produced, but one concern is the way the author put together. Because there were about three states data is not available or included, including how did the government got the unemployment measurement of each states on county level? And if this was involved with human subjects, did the government let them know their data is part of the public data? 
        
    + Population data: 
        + A question is how did the government gathered the data? What was the methods that were used to do so?
        
    + Region data: 
        + How did the government assign each states to certain region? Because it is questionable that WY is consider a part of West Region of the state.
        


